### 2018-10-22 pkg/model4you/inst/JORS/man_model4you.Rnw
revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2353 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 50505e1e661eedd0d20c5ada30d78fc811700cc3

\paragraph{1. Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristicum) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data is partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@
Note that \Sexpr{'pmtree()'} uses the data given in the call of the base
model. It automatically uses variables not used in the model formula
(in the example above \Sexpr{'response ~ treatment'}) as potential subgroup
defining variables. This can be edited using the \Sexpr{'zformula'} argument.

@@

In the following we will present an example application for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed. Also for visualisations we need packages
\emph{ggplot2}, \emph{ggbeeswarm} and \emph{gridExtra}.
<<prep>>=
library("model4you")
set.seed(2017)

@@

\paragraph{Mathematics exam analysis:}
In 2014 first-year business and economics students at the University of
Innsbruck were divided into two examination groups. Group 1 wrote the exam in
the morning and group 2 started after the first group finished. The exams for
the two groups were slightly different.  The data can be accessed and prepared
as follows:

@@

To investigate whether the exam was fair, we asses whether the two groups
differ in terms of the percentage of correctly answered exam questions. This
can be done using a simple linear model regressing the percentage points of
correct anwers on the exam group.
<<math_bmod>>=
bmod_math <- lm(pcorrect ~ group, data = MathExam)
@
The estimtates and confidence intervals of this model can be computed via
<<>>=
cbind(estimate = coef(bmod_math), confint(bmod_math))
@
The model can be visualised by plotting the estimated densities
(see Figure~\ref{fig:math_bmod_vis}):
<<math_bmod_vis, fig.height=2.5, fig.width=5, out.width="0.7\\textwidth", fig.cap = "Density estimates of base model for the Mathematics Exam data.">>=
lm_plot(bmod_math)
@
Both the estimates and confidence intervals and the density curves suggest
that there is almost no difference between the two groups. But does this
really hold for all types of students?

@@

The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \Sexpr{'tests'}. These online tests were
conducted biweekly throughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well throughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).

@@

Dependence plots with the group effect (treatment effect) on the y-axsis and
the student characteristics on the x-axis are a good way of visualising the
personalised models and for getting knowledge about the interactions between
student characteristics and the treatment. Since the percentage of successfull
online tests is measured on a grid, a bee plot possibly shows the relationships
even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1}).
<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Bee plot.")>>=
dpdat_math <- cbind(pmods_math, MathExam)

@@

For the number of previous attempts to pass the exam and the gender box plots,
bee plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=

@@

With the tools provided by the \emph{model4you} package it is
very simple to create understandable stratified and personalised models and
compelling visualisations that can be used to communicate theses models.

>>>>

\paragraph{1. Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristic) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data is partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@
Note that \Sexpr{'pmtree()'} uses the data given in the call of the base
model. It automatically uses variables not used in the model formula
(in the example above \Sexpr{'response ~ treatment'}) as potential subgroup
defining variables. This can be edited using the \Sexpr{'zformula'} argument.

@@

In the following we will present an example application for model-based trees
and personalised models. For this we need to load the package and -- to ensure
reproducibility -- set a random seed.
\new{
Also for visualisations we need packages
\emph{ggplot2} \citep{ggplot2}, \emph{ggbeeswarm} \citep{ggbeeswarm} and
\emph{gridExtra} \citep{gridExtra}. The data used is
available in the \emph{psychotools} package \citep{psychotools}.
}
<<prep>>=
library("model4you")
set.seed(2017)

@@

\paragraph{Mathematics exam analysis:}
In 2014 first-year business and economics students at the University of
Innsbruck were divided into two examination groups. Group 1 took the exam in
the morning and group 2 started after the first group finished. The exams for
the two groups were slightly different.  The data can be accessed and prepared
as follows:

@@

To investigate whether the exam was fair, we asses whether the two groups
differ in terms of the percentage of correctly answered exam questions. This
can be done using a simple linear model regressing the percentage points of
correct anwers on the exam group.
<<math_bmod>>=
bmod_math <- lm(pcorrect ~ group, data = MathExam)
@
The estimates and confidence intervals of this model can be computed via
<<>>=
cbind(estimate = coef(bmod_math), confint(bmod_math))
@
The model can be visualised by plotting the estimated densities
(see Figure~\ref{fig:math_bmod_vis}):
<<math_bmod_vis, fig.height=2.5, fig.width=5, out.width="0.7\\textwidth", fig.cap = "Density estimates of base model for the Mathematics Exam data.">>=
lm_plot(bmod_math)
@
Both the estimates and confidence intervals and the density curves suggest
that there is almost no difference between the two groups. But does this
really hold for all types of students?

@@

The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \Sexpr{'tests'}. These online tests were
conducted biweekly throughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well throughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).
\new{Note that confidence intervals shown in the tree are not adjusted and should
hence be interpreted as a measure for variablity and not be used for inference.}

@@

Dependence plots with the group effect (treatment effect) on the y-axsis and
the student characteristics on the x-axis are a good way of visualising the
personalised models and for getting knowledge about the interactions between
student characteristics and the treatment. Since the percentage of successfull
online tests is measured on a grid, a \new{beeswarm} plot possibly shows the relationships
even better than the scatter plot (both shown in Figure~\ref{fig:math_dp1}).
<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot.", "Beeswarm plot.")>>=
dpdat_math <- cbind(pmods_math, MathExam)

@@

For the number of previous attempts to pass the exam and the gender box plots,
\new{beeswarm} plots or a combination thereof can be used (Figure~\ref{fig:math_dp2}).
<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Number of previous attempts.', 'Gender.')>>=

@@

With the tools provided by the \emph{model4you} package it is
very simple to create understandable stratified and personalised models and
compelling visualisations that can be used to communicate these models.
